---
title: "Supervised learning competition"
author: 
  - Aaron van Riet
  - Max Verwijmeren
  - Nelson Durañona Sosa
date: 2024-10-18
format:
  html:
    toc: true
    self-contained: true
    code-fold: true
    df-print: kable
---


```{r}
#| label: R packages
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

# additional packages here
source("pre-processing_utils.R")
source("analysis_utils.R")
source("plotting_utils.R")
```

```{r}
#| label: data loading
#| echo: false

# your R code to load the data here
train <- readRDS("raw_data/train.rds")
test <- readRDS("raw_data/test.rds")
### Split the original train set into 80% and 20%
train.splits <- train |>
  mutate(split = sample(rep(c("train", "test"), times = c(253, 63))))
train.subset <- train.splits |> filter(split == "train")
test.unseen <- train.splits |> filter(split == "test")
train.numeric <- df_as_numeric(train)

```
 


# Data description

*Describe the data and use a visualization to support your story. (approx. one or two paragraphs)*

The database consists on record students within 360 observations. It was not necessary imputation methods since all values are present, checked with `sum(is.na(train)) == 0 `.  
 
Firstly, we explore the distribution of numerical values and categorical:


```{r}
#| label: eda visualization1
#| warning: false

plots <- plot_histograms_and_bars(train)
print(plots$numeric_plot)
print(plots$categorical_plot)

```



The main goal is the prediction of the score using available attributes. Consequently, we are interested in the relationship between the `score` and other attributes. In the following correlation graph we can notice positive a negative contributions to the score. The most relevant negative correlation is `failures` followed by `gout`, `age`,`schoolsup`,`Mjob_other`. Conversely, the most important positive correlation  `Medu``Fedu` `higher` `studytime`.  




# Model description

*Briefly describe which models you compare to perform prediction. (approx. two or three paragraphs)*

## Linear Regression

In order to , try the easy solution first. We use a multiple linear regression model. 

We applied the following steps: 

1. Initially we started with a simple predictor using `failures` because it has the strongest correlation with the response variable `score`   

2. We applied attribute selection iterating over all other predictors :
  2.1 Cross-validation with k-folds incorporating this tentative predictor  
  2.2 After try all predictors by choose which that minimizes the MSE error
  2.3 Iterate and pick another predictor , repeat the process.
3. We after 7 predictors as we notice the MSE was not improving by adding more.

The final predictors is given by 

```{r}
#| label: eda visualization2
#| warning: false


lm_formula <- 'score~failures + sex+Medu + schoolsup + goout + romantic + Mjob_other'


linearReg.model <- lm(formula = lm_formula, data = train.numeric)
coef(linearReg.model)
```


## Classification trees - Boosting  


## Classification trees - Random Forest 


# Data transformation and pre-processing

*Describe additional pre-processing steps you have used, if any (e.g., dealing with categorical data, scaling). If you do not do any pre-processing, you can leave this section out.*



# Model comparison

Describe how you compare the methods and why. (approx. two or three paragraphs)

# Chosen model

Show which method is best and why. (approx. one paragraph) You are welcome to use tables and plots!

```{r}
#| label: table example
data.frame(
  model       = c("Cool model 1", "Cool model 2"),
  performance = c(1.2, 1.8),
  other       = c(0.5, 0.3),
  notes       = c("Some note", "another note")
)
```

# Team member contributions

Write down what each team member contributed to the project.

- Author One: a, b, c
- Author Two: b, c, d
- Author Three: a, b, d
